---
title: "Simple template for R Markdown"
subtitle: "for Advanced Methods for Regression and Classification"
author: "Prof. Peter Filzmoser"
date: "01.10.2024"
output: pdf_document
---

```{r}
data(College,package="ISLR")
str(College)
```

```{r}
summary(College)
```
Our goal is to find a linear regression model which allows to predict the variable
Apps, i.e. the number of applications received, using the remaining variables except of the
variables Accept and Enroll.

For the following tasks, split the data randomly into training and test data (about
2/3 and 1/3), build the model with the training data, and evaluate the model using the
RMSE as a criterion. 

split the data into training and test data:
```{r}
set.seed(123)
n <- nrow(College)
train <- sample(1:n, n/3)
test <- -train
train.data <- College[train,]
test.data <- College[test,]
```


creating the model
```{r}
model <- lm(Apps ~ . - Accept - Enroll, data=train.data)
summary(model)
```

predict the number of applications for the test data:
```{r}
pred <- predict(model, newdata=test.data)
```

calculate the RMSE:
```{r}
rmse <- sqrt(mean((test.data$Apps - pred)^2))
rmse
```
